---
title: "Visualizations in R"
author: "Alex Slavenko and Maria Novosolov"
output:  
  html_document:  
    toc: true  
    toc_float: true
    smooth_scroll: true
    collapsed: false
    # number_sections: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this workshop we will focus three topics:

1. How to preper our plots for publication

2. How to create interactive visualizations

3. How to create geographic visualization

We will use data on alcohol consumption in the world per person per capita. At first we will explor the data in regular types of plots. And then we will upload again the same data in ashapefile format to learn how to plot the data on a global map.

First thing is to make sure you have all the libraries uploaded for this session

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(leaflet)
library(geojsonio)
library(ggpubr)
library(rvg)
library(ggvis)
```

Next, upload the data into your work environment. Make sure to change the path that you are using in the following code to fit the path for the file in your computer.

```{r}
data.vis<- read.csv("~/Dropbox/Studies/ecodatasci_tlv/visualizations/alcohol-consumption-per-person_VinR.csv")
```
 

# Publication quality plots

## Colour schemes

## Background layers (theme layers)

## Legend manipulation

## Axes manipulation

## Facets, rotations, etc.

## Export

### High quality files

### Export to office

# Interactive visualization

In this part we will learn how to create interactive plots and how to use them online and in presentations or Rmd files for reporting our results.

## Introduction to plotly

Plotly is a grate cross platform package that allows a lot of variation for plotting interactive plots. We will use the R version of the package. Lets start with a simple use of the package and that is transforming our ggplots to interactive visualizations with the code ggplotly()

```{r}
# ggplotly()
```

Now we can also just use 
## Introduction to ggvis

## Introduction to 3D


# Geographic plotting

## ggplot2

## leaflet

Read the map shapefile into R. We use geoJSON because it's a lighter file format for geographical data
```{r}
#Data.vis.shape <- geojsonio::geojson_read("shapefile.json", what = "sp")
```

